### 2011-12-17 doc/theses/mikula/proposed_algorithm.tex
Master thesis: In progress.

\subsection{Syntax Tree Nodes}
\todo[inline]{Dat do appendixu?}
Nodes of the syntax are organized in an is-a hierarchical structure, commonly used in the object oriented programming languages, where a simple node can be of several types. Nodes with emphasised names express a common type of a group of nodes - their descendants, and these type nodes cannot be directly used in the syntax tree. The remaining nodes represent particular constructs of the XQuery language and their instances can be used as nodes of the syntax tree. Each of the instances additionaly contains details about the particular form of the respective XQuery construct used in the query.

@@

For example, an instance of the syntax tree cannot directly contain nodes $Node$ and $ExprNode$, but it can contain instances of nodes $AttributeNode$ and $LiteralNode$. Regarding the types, an instance of $LiteralNode$ node is of three types: $LiteralNode$, $ExprNode$ and $Node$. It carries information on a type of the literal and its value, for example $xs:integer$ of value $5$.

@@

\begin{description}
\item \textbf{Node}
\begin{description}
\item AttributeNode (I)
\item AttrListNode
\item AxisNode (I)
\item CaseClauseNode (I)
\item CaseClausesNode (I)
\item CDataSectionNode
\item ContentNode
\item DefaultCaseNode (I)
\item EntityRefNode
\item \textbf{ExprHolderNode}
  \begin{description}
  \item BindingSequenceNode (I)
  \item DefaultValueNode (I)
  \item ElseExpressionNode (I)
  \item OperandExpressionNode (I)
  \item ReturnClauseNode (I)
  \item TestExpressionNode (I)
  \item ThenExpressionNode (I)
  \item WhereClauseNode (I)
  \end{description}
\item \textbf{ExprNode}
  \begin{description}
  \item CommaOperatorNode (I)
  \item ConstructorNode (I)
  \item ContextItemExprNode
  \item EmptySequenceNode
  \item ExtensionExprNode (I)
  \item FLWORExprNode (I)
  \item FunctionCallNode
  \item IfExprNode (I)
  \item LiteralNode
  \item OperatorNode (I)
  \item OrderedExprNode (I)
  \item PathExprNode (I)
  \item QuantifiedExprNode (I)
  \item TypeswitchExprNode (I)
  \item UnorderedExprNode (I)
  \item ValidateExprNode (I)
  \item VarRefNode
  \end{description}
\item FunctionBodyNode
\item CharRefNode
\item InClausesNode
\item \textbf{ItemTypeNode}
  \begin{description}
  \item AnyItemNode
  \item AtomicTypeNode
  \item KindTestNode
  \item NameTestNode
  \end{description}
\item LocationHintNode
\item LocationHintsNode
\item \textbf{ModuleChildNode}
  \begin{description}
  \item ModuleDeclNode
  \item PrologNode
  \item QueryBodyNode
  \end{description}
\item ModuleNode
\item NameNode
\item OrderByClauseNode
\item OrderSpecNode
\item ParamListNode
\item ParamNode
\item PITargetNode
\item PragmaListNode
\item PragmaNode
\item PredicateListNode
\item \textbf{PrologChildNode}
  \begin{description}
  \item BaseURIDeclNode
  \item BoundarySpaceDeclNode
  \item ConstructionDeclNode
  \item CopyNamespacesDeclNode
  \item DefaultCollationDeclNode
  \item DefaultNamespaceDeclNode
  \item EmptyOrderDeclNode
  \item FunctionDeclNode
  \item \textbf{ImportNode}
    \begin{description}
    \item ModuleImportNode
    \item SchemaImportNode
    \end{description}
  \item NamespaceDeclNode
  \item OptionDeclNode
  \item OrderingModeDeclNode
  \item VarDeclNode
  \end{description}
\item StepExprNode
  \begin{description}
  \item SelfOrDescendantStepNode
  \end{description}
\item StringNode
\item TupleStreamNode
\item TypeNode
\item \textbf{VariableBindingNode}
  \begin{description}
  \item ForClauseNode
  \item InClauseNode
  \item LetClauseNode
  \end{description}
\item VarValueNode
\end{description}
\end{description}

@@

\subsection{Description of the Syntax Tree}
Since we modified only the implementation of the output form, the design of the syntax tree remains consequent to the entire design of the analyses. Each node represents a logical component of a query.

Inner nodes are: $AttributeNode$, $AxisNode$, $CaseClauseNode$, $CaseClausesNode$, $DefaultCaseNode$, $BindingSequenceNode$, $DefaultValueNode$, $ElseExpressionNode$, $OperandExpressionNode$, $ReturnClauseNode$, $TestExpressionNode$, $ThenExpressionNode$, $WhereClauseNode$, $CommaOperatorNode$, $ConstructorNode$, $ExtensionExprNode$, $FLWORExprNode$, $IfExprNode$, $OperatorNode$, $OrderedExprNode$, $PathExprNode$, $QuantifiedExprNode$, $TypeswitchExprNode$, $UnorderedExprNode$, $ValidateExprNode$, $InClausesNode$, $QueryBodyNode$, $ModuleNode$, $OrderByClauseNode$, $OrderSpecNode$, $ParamListNode$, $ParamNode$, $PragmaListNode$, $PredicateListNode$, $FunctionDeclNode$, $ModuleImportNode$, $SchemaImportNode$, $VarDeclNode$, $StepExprNode$, $TupleStreamNode$, $ForClauseNode$, $InClauseNode$, $LetClauseNode$.

Nodes that can be both inner and leaf are: $AttrListNode$, $ContentNode$, $FunctionCallNode$, $FunctionBodyNode$, $KindTestNode$, $LocationHintsNode$, $NameNode$, $PrologNode$, $PITargetNode$, $VarValueNode$.

Leaf nodes are: $CDataSectionNode$, $EntityRefNode$, $ContextItemExprNode$, $EmptySequenceNode$, $VarRefNode$, $PragmaNode$, $BaseURIDeclNode$, $BoundarySpaceDeclNode$, $ConstructionDeclNode$, $CopyNamespacesDeclNode$, $DefaultCollationDeclNode$, $DefaultNamespaceDeclNode$, $EmptyOrderDeclNode$, $LocationHintNode$, $NamespaceDeclNode$, $OptionDeclNode$, $OrderingModeDeclNode$, $SelfOrDescendantStepNode$, $StringNode$, $TypeNode$.

\todo{zistit, ci konstanty nahodou nie su jedine. NIE: CDataSectionNode, ContentNode(ale nemusi byt list), EntityRefNode, ContextItemExprNode, EmptySequenceNode, FunctionCallNode(bez parametrov), VarRefNode, FunctionBodyNode(prazdna funkcia), CharRefNode, AnyItemNode, AtomicTypeNode, KindTestNode(nemusi), NameTestNode, LocationHintNode, ModuleDeclNode, PrologNode, NameNode(nemusi), PITargetNode(nemusi), PragmaNode, BaseURIDeclNode, BoundarySpaceDeclNode, ConstructionDeclNode, CopyNamespacesDeclNode, DefaultCollationDeclNode, DefaultNamespaceDeclNode, EmptyOrderDeclNode, NamespaceDeclNode, OptionDeclNode, OrderingModeDeclNode, SelfOrDescendantStepNode, StringNode, TypeNode, VarValueNode}

\todo[inline]{\^ Toto bude potřeba popsat přesně, ne jen příkladem. Klidně převezměnte z diplomky JS.}

In detail, the syntax tree is a n-ary tree representing syntactic structure of a given query. Every node represents a certain XQuery construct and that is a type of the node. Some types can be divided into more specific subtypes. This is shown \todo{obrazok typov uzlov}.

\todo[inline]{\^ Toto je skoro stejný odstavec ne?}

An important characteric of the syntax tree is related to a definition of local variables and their scope. The representation of a definition of a local variable is $VariableBinding$ node. Nodes of this type contain only two subnodes; a type of the variable and a binding expression. The node and its subtree representing the type do not contain any expressions. The binding expression node is an expression node but the represented expression cannot use the variable which is just being bound. Therefore the scope of the new variable is not the subtree of the $VariableBinding$ node. Assuming \todo{Toto je velmi high-level popis na to, jak málo o syntaktickém stromě a dalších položkách víme.} the post-order numbering of the nodes, the new variable may be valid in some nodes with a higher post-order sequence number. Which \todo{Toto není anglický slovosled.} nodes that are, it depends on the type of the ancestor node(s).

>>>>

\subsection{Syntax Tree Node Types}
\todo[inline]{Dat figures do appendixu?}
The node types (or classes) of the syntax are organized in an is-a hierarchical structure, commonly used in the object oriented programming languages, where a node can be of several types. This hierarchy is shown in figures \ref{FIG_syntax_tree_node_types_part_1}, \ref{FIG_syntax_tree_node_types_part_2}, \ref{FIG_syntax_tree_node_types_part_3}, \ref{FIG_syntax_tree_node_types_part_4}, \ref{FIG_syntax_tree_node_types_part_5}, \ref{FIG_syntax_tree_node_types_part_6}, \ref{FIG_syntax_tree_node_types_part_7}, \ref{FIG_syntax_tree_node_types_part_8}, \ref{FIG_syntax_tree_node_types_part_9}. Nodes with emphasised names express a common type of a group of nodes - their descendants, and these type nodes cannot be directly used in the syntax tree (abstract classes). The remaining nodes represent particular constructs of the XQuery language and their instances can be used as nodes of the syntax tree. Each of the instances additionaly contains details about the particular form of the respective XQuery construct used in the query.

@@

For example, an instance of the syntax tree cannot directly contain instances of nodes $Node$ and $ExprNode$, but it can contain instances of nodes $AttributeNode$ and $LiteralNode$. Regarding the types, an instance of $LiteralNode$ node is of three types: $LiteralNode$, $ExprNode$ and $Node$. It carries information on a type of the literal and its value, for example $xs:integer$ of value $5$.

\todo[inline]{Asi bude potrebne nakreslit pre kazdy node, co presne obsahuje. Alebo staci pre tie, ktore budeme my pouzivat?}

@@

\begin{figure}
\caption{Syntax tree node types part 1.}
\label{FIG_syntax_tree_node_types_part_1}
\begin{description}
\item \textbf{Node}
\begin{description}
\item AttributeNode (I)
\item AttrListNode (IL)
\item AxisNode (I)
\item CaseClauseNode (I)
\item CaseClausesNode (I)
\item CDataSectionNode (L)
\item ContentNode (IL)
\item DefaultCaseNode (I)
\item EntityRefNode (L)
\item \textbf{ExprHolderNode}
\item \textbf{ExprNode}
\item FunctionBodyNode (IL)
\item CharRefNode (L)
\item InClausesNode (I)
\item \textbf{ItemTypeNode}
\item LocationHintNode (L)
\end{description}
\end{description}
\end{figure}

\begin{figure}
\caption{Syntax tree node types part 2.}
\label{FIG_syntax_tree_node_types_part_2}
\begin{description}
\item \textbf{Node}
\begin{description}
\item LocationHintsNode (IL)
\item \textbf{ModuleChildNode}
\item ModuleNode (I)
\item NameNode (IL)
\item OrderByClauseNode (I)
\item OrderSpecNode (I)
\item ParamListNode (I)
\item ParamNode (I)
\item PITargetNode (IL)
\item PragmaListNode (I)
\item PragmaNode (L)
\item PredicateListNode (I)
\item \textbf{PrologChildNode}
\item StepExprNode (I)
\item StringNode (L)
\item TupleStreamNode (I)
\item TypeNode (L)
\item \textbf{VariableBindingNode}
\item VarValueNode (IL)
\end{description}
\end{description}
\end{figure}

\begin{figure}
\caption{Syntax tree node types part 3.}
\label{FIG_syntax_tree_node_types_part_3}
\begin{description}
\item \textbf{ExprHolderNode}
  \begin{description}
  \item BindingSequenceNode (I)
  \item DefaultValueNode (I)
  \item ElseExpressionNode (I)
  \item OperandExpressionNode (I)
  \item ReturnClauseNode (I)
  \item TestExpressionNode (I)
  \item ThenExpressionNode (I)
  \item WhereClauseNode (I)
  \end{description}
\end{description}
\end{figure}

\begin{figure}
\caption{Syntax tree node types part 4.}
\label{FIG_syntax_tree_node_types_part_4}
\begin{description}
\item \textbf{ExprNode}
  \begin{description}
  \item CommaOperatorNode (I)
  \item ConstructorNode (I)
  \item ContextItemExprNode (L)
  \item EmptySequenceNode (L)
  \item ExtensionExprNode (I)
  \item FLWORExprNode (I)
  \item FunctionCallNode (IL)
  \item IfExprNode (I)
  \item LiteralNode (L)
  \item OperatorNode (I)
  \item OrderedExprNode (I)
  \item PathExprNode (I)
  \item QuantifiedExprNode (I)
  \item TypeswitchExprNode (I)
  \item UnorderedExprNode (I)
  \item ValidateExprNode (I)
  \item VarRefNode (L)
  \end{description}
\end{description}
\end{figure}

\begin{figure}
\caption{Syntax tree node types part 5.}
\label{FIG_syntax_tree_node_types_part_5}
\begin{description}
\item \textbf{ItemTypeNode}
  \begin{description}
  \item AnyItemNode (L)
  \item AtomicTypeNode (L)
  \item KindTestNode (IL)
  \item NameTestNode (L)
  \end{description}
\end{description}
\end{figure}

\begin{figure}
\caption{Syntax tree node types part 6.}
\label{FIG_syntax_tree_node_types_part_6}
\begin{description}
\item \textbf{ModuleChildNode}
  \begin{description}
  \item ModuleDeclNode (L)
  \item PrologNode (IL)
  \item QueryBodyNode (I)
  \end{description}
\end{description}
\end{figure}

\begin{figure}
\caption{Syntax tree node types part 7.}
\label{FIG_syntax_tree_node_types_part_7}
\begin{description}
\item \textbf{PrologChildNode}
  \begin{description}
  \item BaseURIDeclNode (L)
  \item BoundarySpaceDeclNode (L)
  \item ConstructionDeclNode (L)
  \item CopyNamespacesDeclNode (L)
  \item DefaultCollationDeclNode (L)
  \item DefaultNamespaceDeclNode (L)
  \item EmptyOrderDeclNode (L)
  \item FunctionDeclNode (I)
  \item \textbf{ImportNode}
    \begin{description}
    \item ModuleImportNode (I)
    \item SchemaImportNode (I)
    \end{description}
  \item NamespaceDeclNode (L)
  \item OptionDeclNode (L)
  \item OrderingModeDeclNode (L)
  \item VarDeclNode (I)
  \end{description}
\end{description}
\end{figure}

\begin{figure}
\caption{Syntax tree node types part 8.}
\label{FIG_syntax_tree_node_types_part_8}
\begin{description}
\item StepExprNode (I)
  \begin{description}
  \item SelfOrDescendantStepNode (L)
  \end{description}
\end{description}
\end{figure}

\begin{figure}
\caption{Syntax tree node types part 9.}
\label{FIG_syntax_tree_node_types_part_9}
\begin{description}
\item \textbf{VariableBindingNode}
  \begin{description}
  \item ForClauseNode (I)
  \item InClauseNode (I)
  \item LetClauseNode (I)
  \end{description}
\end{description}
\end{figure}

@@

\subsection{Description of the Syntax Tree}
An important characteric of the syntax tree is related to a definition of local variables and their scope. The representation of a definition of a local variable is $VariableBinding$ node. Nodes of this type contain only two subnodes; a type of the variable and a binding expression. The node and its subtree representing the type do not contain any expressions. The binding expression node is an expression node but the represented expression cannot use the variable which is just being bound. Therefore the scope of the new variable is not the subtree of the $VariableBinding$ node. Assuming \todo{Toto je velmi high-level popis na to, jak málo o syntaktickém stromě a dalších položkách víme.} the post-order numbering of the nodes, the new variable may be valid in some nodes with a higher post-order sequence number. It depends on the type of the ancestor node(s) which nodes that are.
